---
title: "Inter-related Time-to-Event and Longitudinal Processes"
subtitle: "A causal approach to data generation"
author: "Afshin Mashadi-Hossein"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style type="text/css">
.remark-slide-content {
    font-size: 15px;
}

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

.remark-code, .remark-inline-code { font-family: 'Source Code Pro', 'Lucida Console', Monaco, monospace;
                                    font-size: 90%;
                                  }
</style>

```{r setup, include=F}
source("./R/global.R")
```

## Overview

.pull-left[
### 1- Motivation
 - [Rievew of the problem](#3)
 - [Why simulation and why DAG-based](#4)


### 2- Building the DAG sequentially
 - [Step 1: Expected longitudinal Process](#7)
 - [Step 2: Added the random patient effect](#8)
 - [Step 3: Adding observation noise](#11)
 - [Step 4: Adding some baseline factors](#12)
 - [Step 5: Adding TTE as a function longitudinal process](#13)
]

.pull-right[
### 3- Evaluating the Simulation
 - [Evaluting the TTE outcome](#22)
 - [KM plots](#23)
 - [Cumulative hazard plots](#24)

### 4- Concluding Thoughts
- [Take away, next steps and acknowledgement](#24)

]


---
## Review (JM Intro)

<iframe src="https://amashadihossein.github.io/jm_intro/#3"
height="535" width="720" ></iframe>


---
## Simulating joint TTE/longitudinal process is useful

### Why simulation? 

- Establish qualitative understanding of the dynamics
- Establish rough quantitative understanding of model parameters
- Power analysis in complex setting
- Method comparison

### Why a causal approach?

- Using a causal language of Directed Acyclic Graphs (DAGs) is accessible to SME
 collaborators
- Simulation via DAGs more amenable to causal inference frameworks
- Available JM solvers have limitations by design or in practice (computational)

---
## Two views on data generating approach

### The model-first view
- Simulation as described [here](https://amashadihossein.github.io/jm_intro/#26),
has a probabilistic-model-first view of the data generating mechanism

- It more or less seeks a minimalist set of parameters to describe a single 
joint-likelihood that can do a decent job describing the process

### The process-first view

- Simulation based on **sequential** realization of nodes in a DAGs over time, 
views data generation closer to what naturally/intuitively occurs in the process.

- In practice the number of parameters grow as a function of time

- Getting the data output to conform to a minimalist probabilistic model 
requires more work though

---
class: center, middle
# Building the DAG sequentially

---
## Step 1: The expected longitudinal process

To abstract out the mechanistic modeling of the expected temporal pattern, which
while interesting could be reasonably considered out of scope, We could take the 
"typical" temporal pattern to describe $E[y|t,group]$
<iframe src="https://amashadihossein.github.io/jm_intro/#28"
height="482" width="650" ></iframe>

---
## Step 2: Adding the random effect to unit of obs (patient)

The probabilistic model envisioned a mixed effect to explain deviation of each
data point from $E[y|t=t^*,group=group^*]$


### General:

$$y_i(t) = x_i(t) \beta + z_i(t) b_i + \epsilon_i(t) \; (1)$$
$$\eta_i(t) := E[y_i] = x_i(t) \beta + z_i(t) b_i$$

- $y_i(t)$ is the observed value of biomarker at time $t$
- $\eta_i(t)$ is the unobserved true value of biomarker at time $t$
- $\beta$ is the vector of fix-effect parameters
- $b_i$ is the vector of random-effect parameters for subject $i$ ~ $N(0, D)$
- $\epsilon_i$ is the iid Gaussian noise ~ $N(0,\sigma)$





### Specific implementation:

In the specific case of this simulation, $b_i$ was a random intercept  ~ $N(0, \sigma)$
To translated to DAG



.footnote[NOTE: previous simulation used an AR1 for covariance matrix $D$] 

---

## Step 2: `simcausal` implementation

Data generating process: assignment of group $\rightarrow$ random effect
$\rightarrow$ average temporal profile by group 

```{r dag, message=FALSE,echo=T}
t_obs <- 1:30
t.end <- length(t_obs)
D <- DAG.empty() +
  node(name = "grps", distr = "rbern", prob = .5) +
  node("unitnoise",t=1, distr="rnorm", mean=0, sd=.05) +
  node("ybar", t = 1:t.end, distr = "aveProfile", time=t, 
       grp=grps, unitnoise= unitnoise[1])
obj_D <- set.DAG(D)
```

.pull-left[

#### DAG: shown t = 1, 2 (out of 30) 
```{r, echo=F, message=F, fig.retina = 6, fig.height=5, fig.width= 6, fig.align='center'}
set.seed(9)
simcausal::plotDAG(obj_D,
                   vertex_attrs = list(size = 40, label.cex = 1.5),
                   xjitter = 5, yjitter = 5,
                   tmax = 2)
```

]

.pull-left[

#### Simulation: n = 100

```{r, echo=T, message=F}
sim_ybar <- simcausal::sim(DAG = obj_D,n = 100)
glue::glue("columns:{ c(colnames(sim_ybar)[c(1:4)], \"...\" ,
           colnames(sim_ybar)[33])}")
```

]

---
## Step 2: `simcausal` implementation $\rightarrow$ visulized

```{r, echo=T, message=F,fig.retina = 6, fig.height=5, fig.width= 6, fig.align='center'}
sim_ybar %>% tidyr::pivot_longer(cols = ybar_1:ybar_30,values_to = "ybar") %>%
    mutate(t = as.numeric(gsub("ybar_","",name))) %>%
    mutate(grps = as.character(grps)) %>%
    ggplot(aes(x = t, y=ybar, colour = grps))+
    geom_line(aes(group = ID), alpha = .2)+
    ylab(expression(eta[i](t)))+
    ggsci::scale_color_jama()
  
```

---
## Step 3: Adding observartion noise

```{r, message=FALSE,echo=T}
D <- D + node("y",t = 1:t.end, distr = "rnorm", mean = ybar[t], sd = 0.02)
obj_D <- set.DAG(D)
```

.pull-left[

#### DAG: shown t = 1, 2, 3 (out of 30) 

```{r, echo=T, message=F, fig.retina = 6, fig.height=5, fig.width= 6, fig.align='center'}
simcausal::plotDAG(obj_D,
                   edge_attrs = list(arrow.size = 1),
                   vertex_attrs = list(label.cex = 1.5),
                   tmax = 3)
```

]

.pull-right[

#### Simulation: n = 100

```{r, echo=T, message=F}
sim_y <- simcausal::sim(DAG = obj_D,n = 100)
```

```{r, echo=F, message=F, fig.retina = 6, fig.height=5, fig.width= 6, fig.align='center'}
sim_y %>% 
  select(ID,grps,contains("y_")) %>%
  tidyr::pivot_longer(cols = y_1:y_30, values_to = "y") %>%
  mutate(t = as.numeric(gsub("y_","",name))) %>%
  mutate(grps = as.character(grps)) %>%
  ggplot(aes(x = t, y=y, colour = grps))+
  geom_line(aes(group = ID), alpha = .2)+
  ggsci::scale_color_jama()
```

]

---
## Step 4: Adding some baseline factors

To avoid crowding the DAG too much, we add a single baseline factor, `x1`. This
baseline will factor in probability of event and remains directly unrelated to 
the longitudinal process

```{r, echo=T, message=FALSE}
D <- D + node(name = "x1",t=1, distr = "rnorm", mean = 0, sd = 1/15)
obj_D <- set.DAG(D)
simcausal::plotDAG(obj_D,
                   edge_attrs = list(arrow.size = 1),
                   vertex_attrs = list(label.cex = 1.5),
                   tmax = 3)
```


---
## Step 5: Adding the TTE as function of $\eta_i(t)$

### General:

The link between probability of and event and unobserved value of the biomarker, 
$\eta_i$, was defined through the hazard model eq (2).

$$h_i(t | M_i(t), w_i) :=$$
$$lim_{\Delta \to 0} P\lbrace t \leq T < t + \Delta t | T \geq t, M_i(t), w_i\rbrace/ \Delta t =$$
$$h_0(t)exp\lbrace w_i \gamma + f(\alpha, w_i, b_i, M_i(t))\rbrace \; (2)$$
<br>
Commonly used examples of functions $f$ are
$$f:= \alpha \eta_i(t) \; (3)$$
$$f:= \alpha_1 \eta_i(t) + \alpha_2 \eta_i^{'}(t)\; (4)$$ where
$$\eta_i^{'}(t) = \frac{d\eta_i(t)}{dt}\; (5)$$
---
## Step 5: Adding the TTE as function of $\eta_i(t)$

### Specific implementation:

- Base hazard was defined $h_0(t) = \lambda a t^{a-1}$ with $a = 1.5$ and $\lambda = 1$
- $f$ was defined as eq(3) with $\alpha = 3$ (unless otherwise stated)
- $w_i ~ N(\mu = 0,\sigma = 1/15)$ and $\gamma\ = 1$

To calculate hazard
```{r, echo=TRUE}
add_hz <- function(t, eta, alpha = 3, gamma_x = 0, beta_w = 0,
                  shape_wb = 1.5, scale_wb = 1){
  exp(log(scale_wb) +
        log(shape_wb) + (shape_wb - 1) * log(t) +  
        gamma_x + # baseline covariate
        beta_w + # confounder
        eta * alpha )
}
```



---
## $\bar{\eta}(t) \leftrightarrow \bar{h}(t)$ relation

```{r,echo=T, message=F, fig.retina = 6, fig.height=3, fig.width= 10, fig.align='center'}
hz_bar <- yspline_get() %>%
  data.frame(check.names = F) %>%
  mutate(t = 1:nrow(.)) %>%
  tidyr::pivot_longer(cols = -t,values_to = "eta_bar", names_to = "grps") %>%
  mutate(h_bar = add_hz(t = t/30,eta = eta_bar)) %>% 
  tidyr::pivot_longer(cols = c(eta_bar,h_bar), names_to = "param") 

hz_bar %>% ggplot(aes(x = t, y = value, colour = grps)) +
  geom_line()+
  ylab(NULL)+
  facet_wrap(~ param,scales = "free_y")+
  ggsci::scale_color_jama()
```

.footnote[NOTE: to show `h_bar` trend, $\gamma$ was set to zero for the plots] 
---
## Step 5: Connecting $h_i(t)$ with discrete probabilty of event


With $h(t)$ calculation specified, we can use eq. (5.1) to compute discrete 
probability of event at each time point using numerical integration of $h(t)$.

$$p_{event}(t = t^*) = p(t <= t^*) - p(t < t^*)\;$$
$$=F(t^*) - F(t^* -1)\;$$

$$= 1 - S(t^*) - (1 - S(t^* -1))\;$$

$$= S(t^* - 1) - S(t^*)\;$$

$$= exp(-H(t^* - 1)) - exp(-H(t^*))\;$$
$$= e^{\int_{0}^{t^*-1}h(u)du} - e^{\int_{0}^{t^*}h(u)du}\; (5.1)$$

- $S(t)$ is survival function
- $F(t)$ is the CDF of the event
- $H(t)$ is cumulative hazard at $t$ and equal to $-ln(S(t))$ see [eq(9)](https://amashadihossein.github.io/jm_intro/#35)

---
## Step 5: Building the DAG

Putting it all in a single page:

```{r, message=F, echo=TRUE}
#Generate the simcausal DAG
set.seed(9)
t_obs <- 1:30
t.end <- length(t_obs)
D <- DAG.empty()
D <- D +
  node(name = "w", distr="rbern", prob=0.5) + 
  node(name = "grps", distr = "rbern", prob = ifelse(w == 0, 0.8, 0.2)) +
  node(name = "x1", distr = "rnorm", mean = 0, sd = 1/15) +
  node("unitnoise",t=1, distr="rnorm", mean=0, sd=.05) 
D <- D  +
  node("ybar", t = 1:t.end, distr = "aveProfile",time=t, grp=grps, unitnoise= unitnoise[1])+
  node("y",t = 1:t.end, distr = "rnorm", mean = ybar[t], sd = 0.02)

D <- D + 

  node("event", t = 1, distr="rbern", prob=0, EFU=TRUE) +
  node("event", t = 2:t.end,
       distr = "rbern",
  prob = get_event_p(tobs = 1:t, biomarker_tobs = ybar[1:t],
                     event = event[t-1],
                     w = w, x1 = x1,
                     betaW = 0, betaX1 = 0, alpha = 3,
                     t.end = t.end), EFU = T)

obj_D <- set.DAG(D, vecfun = "get_event_p")

```

---
## Step 5: Completing the DAG $\rightarrow$ Visualization

The DAG cut off at $t_{max} = 3$ (for ease of visualization) 

```{r, message=FALSE, echo=T, fig.retina=T, fig.align='center', fig.width= 10, fig.height=6}
simcausal::plotDAG(obj_D,
                   edge_attrs = list(arrow.size = 1),
                   vertex_attrs = list(label.cex = 1.5),
                   tmax = 3)
```

---
## Step 5: Simulation with the assembled DAG

```{r, echo=FALSE, message=FALSE}
sim_dat <- sim(DAG = obj_D, n = 500, rndseed = 123,wide=F)
```

```{r, echo=FALSE, message=FALSE, fig.retina=T, fig.align='center', fig.height=6, fig.width=10}
tmp_d <- sim_dat %>%
  distinct(ID,grps) %>%
  group_by(grps) %>%
  dplyr::sample_n(size = 50) %>%
  ungroup() %>%
  left_join(x = ., y= sim_dat) %>%
  mutate(grps = as.character(grps)) %>%
  group_by(ID) %>%
  mutate(tlast = max(t)) %>%
  mutate(event = as.character(event)) %>%
  ungroup()

g <- tmp_d %>%
  ggplot(aes(x = t,y = y, group = ID, colour = grps)) + geom_line(alpha = .2)+
  geom_point(data = tmp_d %>% filter(t == tlast), mapping = aes(shape = event))+
  ggsci::scale_color_jama()

plotly::ggplotly(g)
```


---
class: center, middle

# Evaluating the simulation

---

## Evaluating the TTE outcome

```{r, echo=TRUE, message=FALSE, results='hide'}
D_a6 <- D + node("event", t = 2:t.end, distr = "rbern",
                 prob = get_event_p(tobs = 1:t, biomarker_tobs = ybar[1:t],
                                    event = event[t-1], w = w, x1 = x1, 
                                    betaW = 0, betaX1 = 0, alpha = 6, 
                                    t.end = t.end), EFU = T)

obj_D_a6 <- set.DAG(D_a6, vecfun = "get_event_p")
sim_dat_a6 <- sim(DAG = obj_D_a6, n = 500, rndseed = 123,wide=F)
```



```{r, echo=FALSE, message=FALSE, fig.retina=T, fig.align='center', fig.height=4, fig.width=8}
tmp_d <- sim_dat_a6 %>%
  distinct(ID,grps) %>%
  group_by(grps) %>%
  dplyr::sample_n(size = 50) %>%
  ungroup() %>%
  left_join(x = ., y= sim_dat) %>%
  mutate(grps = as.character(grps)) %>%
  group_by(ID) %>%
  mutate(tlast = max(t)) %>%
  mutate(event = as.character(event)) %>%
  ungroup()

g <- tmp_d %>%
  ggplot(aes(x = t,y = y, group = ID, colour = grps)) + geom_line(alpha = .2)+
  geom_point(data = tmp_d %>% filter(t == tlast), mapping = aes(shape = event))+
  ggsci::scale_color_jama()

plotly::ggplotly(g)
```

---

## KM plots
.pull-left[  

#### $\alpha = 3$

```{r,echo=T, message=FALSE, fig.retina=T, fig.align='center', fig.height=4, fig.width=4}
sim_dat %>% group_by(ID) %>% mutate(tmax = max(t)) %>% filter(t == tmax) %>%
  ungroup() %>% mutate(t = as.numeric(t)) ->srv_d
survminer::ggsurvplot(survfit(Surv(time = t,event = event ) ~ grps, data=srv_d),
                      pval = T)$plot + ggsci::scale_color_jama()
```
]

.pull-right[

#### $\alpha = 6$
```{r,echo=T, message=FALSE, fig.retina=T, fig.align='center', fig.height=4, fig.width=4}
sim_dat_a6 %>% group_by(ID) %>% mutate(tmax = max(t)) %>% filter(t == tmax) %>%
  ungroup() %>% mutate(t = as.numeric(t)) ->srv_d_a6
survminer::ggsurvplot(survfit(Surv(time = t,event = event ) ~ grps, data=srv_d_a6),
                      pval = T)$plot + ggsci::scale_color_jama()
```
]

---
## Cumulative hazard plots
.pull-left[  

#### $\alpha = 3$

```{r,echo=T, message=FALSE, fig.retina=T, results='hide', fig.keep='all', fig.align='center', fig.height=4, fig.width=4}
ggsurvplot(survfit(Surv(time = t,event = event ) ~ grps, data=srv_d),
           fun = "cumhaz")$plot + ggsci::scale_color_jama()

```
]

.pull-right[

#### $\alpha = 6$
```{r,echo=T, message=FALSE, fig.retina=T, results='hide', fig.keep='all', fig.align='center', fig.height=4, fig.width=4}
ggsurvplot(survfit(Surv(time = t,event = event ) ~ grps, data=srv_d_a6),
           fun = "cumhaz")$plot + ggsci::scale_color_jama()
```
]

---

## Take away, next steps and acknowledgement

- We can simulate the data of JM flavor with DAG-based causal approach fairly well
- `simcausal` provides a good amount utility to make setting up complex DAGs feasible
- Features such as treatment assignment, and flexible effect size estimation via
selection of targets can greatly assist in exploration of the causal dynamics of
the data

### Acknowledgments

- Fred Gruber and Brandon Nathasingh for providing assisting with the set up of
the simulation in `simcausal`
